disaggregation: An \proglang{R} package for Bayesian spatial disaggregation modelling

### 2019-11-22 article.Rnw

Disaggregation modelling, which involves predicting at fine-scale resolution using areal data over heterogenous regions, has become widely used in fields such as epidemiology and ecology. The \pkg{disaggregation} package implements Bayesian spatial disaggregation modelling with a simple, easy to use \proglang{R} interface. The package includes simple data preparation, fitting and prediction functions that allow some user defined model flexibility. In this paper we have presented an application of the package, predicting malaria incidence rate across Madagascar from aggregated count data and environmental covariates.

The modelling framework is implemented using the Laplace approximation and automatic differentiation within the \pkg{TMB} package. This allows fast, optimised calculations in C++. These disaggregation models are computationally intensive and take a long time using Markov Chian Monte Carlo (MCMC) optimisation techniques. Using \pkg{TMB}, the models are much faster and produce similar results

The \pkg{disaggregation} package can be extended to include spatio-temporal disaggregation models. This would require a spatio-temporal field as well as dynamic covariates. However, this would be significantly more computationally intensive. Additionally, tools for cross-validation could be included within the package. Cross-validation of spatial models is non-trivial due to the spatial autocorrelation in the data.

>>>>

Disaggregation modelling, which involves fitting models at fine-scale resolution using areal data over heterogenous regions, has become widely used in fields such as epidemiology and ecology. The \pkg{disaggregation} package implements Bayesian spatial disaggregation modelling with a simple, easy to use \proglang{R} interface. The package includes simple data preparation, fitting and prediction functions that allow some user defined model flexibility. In this paper we have presented an application of the package, predicting malaria incidence rate across Madagascar from aggregated count data and environmental covariates.

The modelling framework is implemented using the Laplace approximation and automatic differentiation within the \pkg{TMB} package. This allows fast, optimised calculations in C++. These disaggregation models are computationally intensive and take a long time using Markov Chian Monte Carlo (MCMC) optimisation techniques. Using \pkg{TMB}, the models are much faster and produce similar results.

Future work will entail extending the \pkg{disaggregation} package to include spatio-temporal disaggregation models. This would require a spatio-temporal field as well as dynamic covariates. However, this would be significantly more computationally intensive. Additionally, tools for cross-validation could be included within the package. Cross-validation of spatial models is non-trivial due to the spatial autocorrelation in the data.
